Gaussian Processes Autoregressive Models for the Disturbance Storm Time Index

### 2016-04-24 swsc.tex
Updates from ShareLaTeX

The simplest way to perform \emph{model selection} is called \emph{grid search} which entails evaluating the above criteria on a fixed grid which is constructed by making a list of values for each hyper-parameter and taking the combinations of all such values. There are more sophisticated methods for \emph{model tuning} such as \emph{coupled simulated annealing} (\citet{Xavier-De-Souza2010}) and \emph{simplex search} (\citet{Nelder1965}) but they are computationally more expensive. In this study we employ \emph{grid search} to find the best performing set of hyper-parameter assignments inside a predefined grid of possible hyper-parameter values.

@@

The time lags chosen for \emph{GP-AR} are $p = 6$ while for \emph{GP-ARX} $p=6$, $p_{ex} = 1$ are arrived by experimenting with values starting from one and choosing the value of $p$ yielding best performance on the 63 events in the bench mark.

@@

\subsection{Storm Events}
\citet{Ji2012}} compare a six OSA $D_{st}$ models in their paper by compiling a list of 63 geomagnetic storm events of varying strengths which occurred from 1998-2006. This serves as an important stepping stone for a systematic comparison of existing and new prediction techniques because performance metrics are averaged over a large number of storm events that occurred over a 8 year period.

@@

>>>>

The simplest way to perform \emph{model selection} is called \emph{grid search} which entails evaluating the above criteria on a fixed grid which is constructed by making a list of values for each hyper-parameter and taking the combinations of all such values. There are more sophisticated methods for \emph{model tuning} such as \emph{coupled simulated annealing} (\citet{Xavier-De-Souza2010}) and \emph{simplex search} (\citet{Nelder1965}) but they are computationally more expensive.

In this study we employ \emph{grid search} to find the best performing set of hyper-parameter assignments inside a predefined grid of possible hyper-parameter values.

@@

Model selection of \emph{GP-AR} and \emph{GP-ARX} is performed by grid search methodology and the root mean square error (\emph{RMSE}) on the validation set is used to select the values of the model hyper-parameters. The time lags chosen for \emph{GP-AR} are $p = 6$ while for \emph{GP-ARX} $p=6$, $p_{ex} = 1$ are arrived by experimenting with values starting from one and choosing the value of $p$ yielding best performance on the 63 events in the bench mark.

@@

\citet{Ji2012} compare a six OSA $D_{st}$ models in their paper by compiling a list of 63 geomagnetic storm events of varying strengths which occurred from 1998-2006. This serves as an important stepping stone for a systematic comparison of existing and new prediction techniques because performance metrics are averaged over a large number of storm events that occurred over a 8 year period. They compare the averaged performance of these models on four key performance metrics.

\begin{enumerate}
    \item $RMSE$: The root mean square error
    \item $CC$: Correlation coefficient between the predicted and actual value of $D_{st}$
    \item $\Delta Dst_{min}$: The error in prediction of the peak value of $D_{st}$ for a storm event.
    \item $|\Delta t_{peak}|$: The error in predicting the timing of a storm peak, also called \emph{timing error}. 
\end{enumerate}

Apart from the metrics above, we also generate the empirical distribution of absolute relative errors $|\frac{D_{st} - \hat{D_{st}}}{D_{st}}|$ for the \emph{GP-AR}, \emph{GP-ARX}, \emph{NM} and \emph{TL} models. These empirical cumulative distributions enable us to upper bound the percentage error in $D_{st}$ prediction and attach probability estimates for said bounds.

@@

Table \ref{table:DstModels} gives a brief enumeration of the models compared, we use the performance benchmark results published in \citet{Ji2012} and to them add results obtained from testing \emph{GP-AR}, \emph{GP-ARX} as well as the \emph{Persistence} model on the same list of storms.

For the purpose of generating the cumulative probability plots of the model errors, we must obtain hourly predictions for all the storm events using the \emph{NARMAX} and \emph{TL} models. In the case of the \emph{NM} model, the formula outlined in \citet{balikhin:narmax} is used to generate hourly predictions. For the \emph{TL} model we use real time predictions listed on their website corresponding to all the storm events. The real time predictions are at a frequency of ten minutes, hourly predictions are generated by averaging the $D_{st}$ predictions for each hour.
